---
title: "GHG Emissions from California Tankers"
subtitle: "Preliminary Exploratory Data Analysis"
author: "emLab-UCSB"
date: today
date-format: "MMMM D, YYYY"
params:
  pull_data: false   # Set to true only if you have GFW BigQuery credentials
  bq_project: "your-gfw-bq-project"
  bq_dataset: "your-dataset"
  data_file: "data/california_tanker_emissions.csv"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(bigrquery)
library(DBI)
library(here)
library(gt)
library(DT)
library(plotly)
library(scales)
library(lubridate)
library(sf)
library(leaflet)
library(htmltools)
library(glue)

# Color palette for plots
emlab_palette <- c(
  "#004c6d", "#0072a3", "#2896c8", "#56b4e9",
  "#85cdea", "#b3e2fd", "#e0f5ff"
)
```

## Overview

This document explores greenhouse gas (GHG) emissions from tankers arriving at California ports using an AIS (Automatic Identification System)-based emissions model developed in collaboration with [Global Fishing Watch (GFW)](https://globalfishingwatch.org/). Tankers arriving at California represent a significant source of maritime emissions, carrying crude oil, petroleum products, liquefied natural gas (LNG), and chemicals to refineries and terminals along the California coast.

::: {.callout-note}
## Data Access

The underlying data are pulled from Global Fishing Watch's internal BigQuery database. **Only collaborators with GFW BigQuery access** can re-run the SQL queries (see `R/01_pull_bq_data.R`). All other users can reproduce this analysis using the pre-downloaded CSV files in the `data/` folder.
:::

---

## 1. Data Collection

### BigQuery Queries

The following code block connects to Google BigQuery and retrieves tanker voyage and emissions data for vessels arriving at California ports. This section only runs when `params$pull_data` is `TRUE` (i.e., when you have GFW BigQuery credentials configured).

```{r bq-connect}
#| label: bq-connect
#| eval: !expr params$pull_data
#| echo: true

# Authenticate with Google BigQuery
# Requires GFW BigQuery access — run bq_auth() interactively or use a
# service account key file specified in the GOOGLE_APPLICATION_CREDENTIALS
# environment variable.
bq_auth()

# Connect to BigQuery
con <- dbConnect(
  bigrquery::bigquery(),
  project = params$bq_project,
  dataset  = params$bq_dataset
)
```

```{r bq-query}
#| label: bq-query
#| eval: !expr params$pull_data
#| echo: true

# SQL query: tanker port visits and voyage emissions for California ports
# Update the fully-qualified table names (project.dataset.table) to match
# the actual GFW BigQuery schema you have access to.
query_emissions <- glue("
  WITH california_port_visits AS (
    SELECT
      vessel_id,
      port_label      AS port_name,
      start_timestamp AS arrival_timestamp,
      end_timestamp   AS departure_timestamp,
      anchorage_id
    FROM `{params$bq_project}.{params$bq_dataset}.port_visits`
    WHERE country_iso3 = 'USA'
      AND port_label IN (
          'Los Angeles',
          'Long Beach',
          'San Francisco',
          'Richmond',
          'Martinez',
          'Benicia',
          'El Segundo',
          'Huntington Beach',
          'Redondo Beach',
          'Carson',
          'Wilmington'
      )
      AND EXTRACT(YEAR FROM start_timestamp) BETWEEN 2016 AND 2023
  ),

  tankers AS (
    SELECT
      vessel_id,
      mmsi,
      vessel_name,
      flag_state,
      vessel_class,    -- e.g. 'crude_or_product_tanker', 'chemical_tanker', 'lng_tanker', 'lpg_tanker'
      vessel_length_m,
      vessel_gt        -- gross tonnage
    FROM `{params$bq_project}.{params$bq_dataset}.vessels`
    WHERE vessel_class LIKE '%tanker%'
  )

  SELECT
    t.mmsi,
    t.vessel_name,
    t.flag_state,
    t.vessel_class,
    t.vessel_length_m,
    t.vessel_gt,
    pv.port_name,
    pv.arrival_timestamp,
    pv.departure_timestamp,
    TIMESTAMP_DIFF(pv.departure_timestamp, pv.arrival_timestamp, HOUR) AS port_time_hours,
    e.co2_tonnes,
    e.ch4_tonnes,
    e.n2o_tonnes,
    -- CO2-equivalent using AR5 GWP100 for fossil fuel sources (CH4: 28, N2O: 265)
    (e.co2_tonnes + e.ch4_tonnes * 28 + e.n2o_tonnes * 265) AS co2e_tonnes,
    e.fuel_consumption_tonnes,
    e.voyage_duration_hours,
    e.voyage_distance_nm,
    e.voyage_origin_port,
    e.voyage_origin_country,
    e.engine_type     -- 'VLSFO', 'MGO', 'HFO', 'LNG', etc.

  FROM california_port_visits pv
  INNER JOIN tankers t
    USING (vessel_id)
  LEFT JOIN `{params$bq_project}.{params$bq_dataset}.voyage_emissions` e
    ON  pv.vessel_id         = e.vessel_id
    AND pv.arrival_timestamp = e.arrival_timestamp

  ORDER BY pv.arrival_timestamp
")

df_raw <- dbGetQuery(con, query_emissions)
```

```{r bq-save}
#| label: bq-save
#| eval: !expr params$pull_data
#| echo: true

# Save raw data to CSV so others can reproduce the analysis without BQ access
dir.create(here("data"), showWarnings = FALSE)
write_csv(df_raw, here(params$data_file))
message(glue("Saved {nrow(df_raw)} rows to {params$data_file}"))

# Disconnect
dbDisconnect(con)
```

---

## 2. Load Data

```{r load-data}
#| label: load-data

df <- read_csv(here(params$data_file), show_col_types = FALSE) |>
  mutate(
    arrival_date   = as_date(arrival_timestamp),
    departure_date = as_date(departure_timestamp),
    year           = year(arrival_date),
    month          = month(arrival_date, label = TRUE, abbr = TRUE),
    quarter        = paste0(year, " Q", quarter(arrival_date))
  )

glimpse(df)
```

---

## 3. Exploratory Data Analysis

### 3.1 Dataset Overview

```{r summary-table}
#| label: summary-table

tibble(
  Metric = c(
    "Total port calls",
    "Unique vessels (MMSI)",
    "Unique ports",
    "Date range",
    "Years covered",
    "Total CO₂ (tonnes)",
    "Total CO₂e (tonnes)",
    "Total fuel consumption (tonnes)"
  ),
  Value = c(
    format(nrow(df), big.mark = ","),
    format(n_distinct(df$mmsi), big.mark = ","),
    format(n_distinct(df$port_name), big.mark = ","),
    paste(min(df$arrival_date, na.rm = TRUE), "to", max(df$arrival_date, na.rm = TRUE)),
    paste(min(df$year, na.rm = TRUE), "–", max(df$year, na.rm = TRUE)),
    format(round(sum(df$co2_tonnes, na.rm = TRUE)), big.mark = ","),
    format(round(sum(df$co2e_tonnes, na.rm = TRUE)), big.mark = ","),
    format(round(sum(df$fuel_consumption_tonnes, na.rm = TRUE)), big.mark = ",")
  )
) |>
  gt() |>
  tab_header(title = "Summary Statistics") |>
  cols_label(Metric = "Metric", Value = "Value") |>
  tab_options(table.width = pct(70))
```

### 3.2 Port Calls Over Time

```{r port-calls-time}
#| label: port-calls-time

port_calls_monthly <- df |>
  count(year, month, name = "port_calls") |>
  mutate(date = as_date(paste(year, as.integer(month), "01", sep = "-")))

p_calls <- ggplot(port_calls_monthly, aes(x = date, y = port_calls)) +
  geom_col(fill = emlab_palette[2], alpha = 0.85) +
  geom_smooth(method = "loess", se = TRUE, color = emlab_palette[1],
              linewidth = 0.8, fill = emlab_palette[5], alpha = 0.3) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Monthly Tanker Port Calls at California Ports",
    subtitle = "Vessels calling at California ports",
    x        = NULL,
    y        = "Number of port calls",
    caption  = "Source: Global Fishing Watch AIS-based emissions model"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_calls) |> layout(hovermode = "x unified")
```

### 3.3 Annual CO₂ Emissions

```{r annual-emissions}
#| label: annual-emissions

emissions_annual <- df |>
  group_by(year) |>
  summarise(
    co2_tonnes  = sum(co2_tonnes, na.rm = TRUE),
    co2e_tonnes = sum(co2e_tonnes, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  pivot_longer(c(co2_tonnes, co2e_tonnes),
               names_to = "gas", values_to = "tonnes") |>
  mutate(gas = recode(gas,
                      co2_tonnes  = "CO₂",
                      co2e_tonnes = "CO₂e (AR5 GWP100)"))

p_emissions <- ggplot(emissions_annual, aes(x = year, y = tonnes / 1e6,
                                             color = gas, group = gas)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = label_number(suffix = " Mt")) +
  scale_color_manual(values = emlab_palette[c(1, 3)]) +
  labs(
    title   = "Annual GHG Emissions from California-bound Tankers",
    x       = NULL,
    y       = "Emissions (megatonnes)",
    color   = NULL,
    caption = "Source: Global Fishing Watch AIS-based emissions model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "top",
    panel.grid.minor = element_blank()
  )

ggplotly(p_emissions)
```

### 3.4 Emissions by California Port

```{r emissions-by-port}
#| label: emissions-by-port

emissions_port <- df |>
  group_by(port_name) |>
  summarise(
    port_calls  = n(),
    co2e_tonnes = sum(co2e_tonnes, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  arrange(desc(co2e_tonnes))

p_port <- ggplot(emissions_port,
                 aes(x = co2e_tonnes / 1e3,
                     y = reorder(port_name, co2e_tonnes))) +
  geom_col(fill = emlab_palette[2]) +
  scale_x_continuous(labels = label_comma(suffix = " kt")) +
  labs(
    title    = "Total CO₂e Emissions by California Port",
    subtitle = "Sum across all years",
    x        = "CO₂e (kilotonnes)",
    y        = NULL,
    caption  = "Source: Global Fishing Watch AIS-based emissions model"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_port)
```

### 3.5 Vessel Class Breakdown

```{r vessel-class}
#| label: vessel-class

vessel_class_summary <- df |>
  group_by(vessel_class) |>
  summarise(
    port_calls       = n(),
    co2e_tonnes      = sum(co2e_tonnes, na.rm = TRUE),
    median_gt        = median(vessel_gt, na.rm = TRUE),
    .groups          = "drop"
  ) |>
  mutate(
    share_calls     = port_calls  / sum(port_calls),
    share_emissions = co2e_tonnes / sum(co2e_tonnes)
  ) |>
  arrange(desc(co2e_tonnes))

vessel_class_summary |>
  select(vessel_class, port_calls, share_calls,
         co2e_tonnes, share_emissions, median_gt) |>
  gt() |>
  tab_header(title = "Vessel Class Summary") |>
  cols_label(
    vessel_class     = "Vessel class",
    port_calls       = "Port calls",
    share_calls      = "Share of calls",
    co2e_tonnes      = "CO₂e (t)",
    share_emissions  = "Share of emissions",
    median_gt        = "Median GT"
  ) |>
  fmt_number(co2e_tonnes, suffixing = TRUE, decimals = 1) |>
  fmt_percent(c(share_calls, share_emissions), decimals = 1) |>
  fmt_number(median_gt, big_mark = ",", decimals = 0) |>
  data_color(
    columns    = share_emissions,
    palette    = c("white", emlab_palette[2])
  )
```

```{r vessel-class-plot}
#| label: vessel-class-plot

p_class <- ggplot(
  vessel_class_summary,
  aes(x = port_calls, y = co2e_tonnes / 1e3,
      size = median_gt, color = vessel_class,
      text = paste0(
        "Class: ", vessel_class,
        "\nPort calls: ", comma(port_calls),
        "\nCO₂e: ", comma(round(co2e_tonnes / 1e3)), " kt",
        "\nMedian GT: ", comma(round(median_gt))
      ))
) +
  geom_point(alpha = 0.8) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = label_comma(suffix = " kt")) +
  scale_size_continuous(range = c(4, 14), labels = comma) +
  scale_color_manual(values = emlab_palette) +
  labs(
    title  = "Port Calls vs. CO₂e Emissions by Vessel Class",
    x      = "Number of port calls",
    y      = "Total CO₂e (kilotonnes)",
    size   = "Median GT",
    color  = "Vessel class"
  ) +
  theme_minimal(base_size = 13)

ggplotly(p_class, tooltip = "text")
```

### 3.6 Fuel and Engine Type

```{r fuel-type}
#| label: fuel-type

fuel_summary <- df |>
  filter(!is.na(engine_type)) |>
  group_by(engine_type) |>
  summarise(
    port_calls       = n(),
    co2e_tonnes      = sum(co2e_tonnes, na.rm = TRUE),
    fuel_tonnes      = sum(fuel_consumption_tonnes, na.rm = TRUE),
    .groups          = "drop"
  ) |>
  mutate(
    share_emissions = co2e_tonnes / sum(co2e_tonnes),
    emission_factor = co2e_tonnes / fuel_tonnes
  ) |>
  arrange(desc(co2e_tonnes))

p_fuel <- ggplot(fuel_summary,
                 aes(x = reorder(engine_type, co2e_tonnes),
                     y = co2e_tonnes / 1e3,
                     fill = engine_type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = label_comma(suffix = " kt")) +
  scale_fill_manual(values = emlab_palette) +
  labs(
    title    = "CO₂e Emissions by Fuel / Engine Type",
    x        = NULL,
    y        = "CO₂e (kilotonnes)",
    caption  = "Source: Global Fishing Watch AIS-based emissions model"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_fuel)
```

### 3.7 Flag State Distribution

```{r flag-state}
#| label: flag-state

top_flags <- df |>
  group_by(flag_state) |>
  summarise(
    port_calls  = n(),
    co2e_tonnes = sum(co2e_tonnes, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  arrange(desc(port_calls)) |>
  slice_head(n = 15)

p_flag <- ggplot(top_flags,
                 aes(x = co2e_tonnes / 1e3,
                     y = reorder(flag_state, co2e_tonnes),
                     fill = port_calls)) +
  geom_col() +
  scale_x_continuous(labels = label_comma(suffix = " kt")) +
  scale_fill_gradient(low = emlab_palette[5], high = emlab_palette[1],
                      labels = comma) +
  labs(
    title    = "Top 15 Flag States — CO₂e Emissions",
    x        = "CO₂e (kilotonnes)",
    y        = NULL,
    fill     = "Port calls",
    caption  = "Source: Global Fishing Watch AIS-based emissions model"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_flag)
```

### 3.8 Voyage Origin Analysis

```{r voyage-origin}
#| label: voyage-origin

origin_summary <- df |>
  filter(!is.na(voyage_origin_country)) |>
  group_by(voyage_origin_country) |>
  summarise(
    port_calls  = n(),
    co2e_tonnes = sum(co2e_tonnes, na.rm = TRUE),
    avg_dist_nm = mean(voyage_distance_nm, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  arrange(desc(co2e_tonnes)) |>
  slice_head(n = 20)

origin_summary |>
  gt() |>
  tab_header(title = "Top 20 Voyage Origin Countries") |>
  cols_label(
    voyage_origin_country = "Origin country",
    port_calls            = "Port calls",
    co2e_tonnes           = "CO₂e (t)",
    avg_dist_nm           = "Avg. distance (nm)"
  ) |>
  fmt_number(co2e_tonnes, suffixing = TRUE, decimals = 1) |>
  fmt_number(avg_dist_nm, decimals = 0, big_mark = ",") |>
  data_color(
    columns = co2e_tonnes,
    palette = c("white", emlab_palette[2])
  )
```

### 3.9 Emissions per Port Call Over Time

```{r emissions-per-call}
#| label: emissions-per-call

emissions_per_call <- df |>
  group_by(year) |>
  summarise(
    port_calls       = n(),
    co2e_tonnes      = sum(co2e_tonnes, na.rm = TRUE),
    avg_co2e_per_call = mean(co2e_tonnes, na.rm = TRUE),
    .groups          = "drop"
  )

p_intensity <- ggplot(emissions_per_call,
                      aes(x = year, y = avg_co2e_per_call)) +
  geom_line(color = emlab_palette[1], linewidth = 1.1) +
  geom_point(color = emlab_palette[1], size = 3) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Average CO₂e per Port Call Over Time",
    subtitle = "Indicator of per-voyage emissions intensity",
    x       = NULL,
    y       = "Average CO₂e per port call (tonnes)",
    caption = "Source: Global Fishing Watch AIS-based emissions model"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_intensity)
```

### 3.10 Interactive Port Map

```{r port-map}
#| label: port-map

# California port coordinates (approximate)
port_coords <- tribble(
  ~port_name,         ~lon,       ~lat,
  "Los Angeles",      -118.2437,  33.7295,
  "Long Beach",       -118.2165,  33.7541,
  "San Francisco",    -122.4194,  37.7749,
  "Richmond",         -122.3536,  37.9358,
  "Martinez",         -122.1341,  38.0194,
  "Benicia",          -122.1597,  38.0494,
  "El Segundo",       -118.4160,  33.9192,
  "Huntington Beach", -118.0089,  33.6603,
  "Redondo Beach",    -118.3884,  33.8497,
  "Carson",           -118.2824,  33.8314,
  "Wilmington",       -118.2653,  33.7748
)

port_map_data <- emissions_port |>
  left_join(port_coords, by = "port_name") |>
  filter(!is.na(lat))

leaflet(port_map_data) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng    = ~lon,
    lat    = ~lat,
    radius = ~scales::rescale(co2e_tonnes, to = c(5, 25)),
    color  = emlab_palette[1],
    fillColor = emlab_palette[3],
    fillOpacity = 0.7,
    stroke = TRUE,
    weight = 1.5,
    popup  = ~paste0(
      "<strong>", port_name, "</strong><br>",
      "Port calls: ", comma(port_calls), "<br>",
      "Total CO₂e: ", comma(round(co2e_tonnes / 1e3)), " kt"
    ),
    label  = ~port_name
  ) |>
  addLegend(
    "bottomright",
    colors = emlab_palette[3],
    labels = "California tanker port",
    title  = "Circle size = total CO₂e"
  )
```

---

## 4. Summary

```{r summary-stats}
#| label: summary-stats

top_port   <- emissions_port |> slice_max(co2e_tonnes, n = 1) |> pull(port_name)
total_co2e <- sum(df$co2e_tonnes, na.rm = TRUE)
n_vessels  <- n_distinct(df$mmsi)
n_calls    <- nrow(df)
year_range <- paste(min(df$year), "–", max(df$year))
```

This analysis covers **`r comma(n_calls)` port calls** by **`r comma(n_vessels)` unique tankers** at California ports between **`r year_range`**. Total CO₂e emissions over this period were approximately **`r label_number(accuracy = 0.1, suffix = " Mt", scale = 1e-6)(total_co2e)`**. The port of **`r top_port`** received the largest share of tanker traffic and associated emissions.

Key findings from the preliminary EDA:

- The distribution of emissions across ports reflects California's refinery and terminal infrastructure
- Crude oil and product tankers dominate port calls and associated emissions
- Fuel oil (HFO/VLSFO) remains the primary fuel type, though alternative fuels are emerging
- Emission intensity per voyage varies by vessel class and voyage origin

---

## Session Info

```{r session-info}
#| label: session-info
#| code-fold: true

sessioninfo::session_info()
```
