---
title: "GHG Emissions from Tankers that Arrive at California Ports"
subtitle: "Preliminary Exploratory Data Analysis"
author: "Environmental Markets Lab (emLab), UCSB"
date: today
date-format: "MMMM D, YYYY"
params:
  pull_data: false   # Set to true only if you have Global Fishing Watch BigQuery credentials
  data_file: "data/tanker_trip_emissions_ending_in_california.csv"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(countrycode)
```

## Overview

This document explores greenhouse gas (GHG) emissions from tankers arriving at California ports using an AIS (Automatic Identification System)-based emissions model developed in collaboration between [Environmental Markets Lab (emLab)](https://emlab.ucsb.edu/) and [Global Fishing Watch (GFW)](https://globalfishingwatch.org/). The model and methodology is described extensively [here](https://emlab-ucsb.github.io/ocean-ghg/).

Tankers arriving at California could be a significant source of maritime emissions, carrying crude oil, petroleum products, liquefied natural gas (LNG), and chemicals to refineries and terminals along the California coast. This report provides a preliminary exploratory data analysis (EDA) of tanker emissions based on the GFW model, covering vessel-level emissions estimates for voyages that end at California ports.

::: {.callout-note}
## Data Access

The underlying data are pulled from Google BigQuery, where emLab and Global Fishing Watch store the vessel-level emissions model. **Only collaborators with GFW BigQuery access** can re-run the SQL queries (see `R/01_pull_bq_data.R`). All other users can reproduce this analysis using the pre-downloaded CSV files in the `data/` folder.
:::

---

## 1. Data Preperation


The following code block connects to Google BigQuery and retrieves tanker voyage and emissions data for vessels arriving at California ports. This section only runs when `params$pull_data` is `TRUE` (i.e., when you have Global Fishing Watch BigQuery credentials configured). This chunk downloads the data and saves it as a CSV file in the `data/` folder, which is then loaded in the next section for analysis. If you do not have BigQuery access, you can skip this section and use the pre-downloaded CSV file.



```{r}
#| label: bq-download data
#| eval: !expr params$pull_data
#| echo: true
# Code to pull data from BigQuery and save as .csv file
# Note: This can only be run by users with access to the BigQuery project and dataset
billing_project <- "emlab-gcp"

# Pull data describing all tanker trips that end in California,
# along with their associated emissions, distance traveled, hours, starting and ending ports, etc
bigrquery::bq_project_query(
  billing_project,
  here::here(
    "sql",
    "tanker_trip_emissions_ending_in_california.sql"
  ) |>
    readr::read_file()
) |>
  bigrquery::bq_table_download(n_max = Inf) |>
  readr::write_csv(here::here(
    params$data_file
  ))
```

## 2. Load Data

Here we load the tanker trip emissions data from the CSV file into R for analysis. This dataset includes vessel-level emissions estimates for voyages that end at California ports, along with associated metadata such as distance traveled, hours at sea, starting and ending ports, and more. The dataset can be directly accessed in the GitHub repo: `r {params$data_file}`.

```{r}
#| label: load-data

tanker_trip_emissions_ending_in_california <- readr::read_csv(
  here::here(params$data_file)
)
```

---

## 3. Exploratory Data Analysis

In this section, we conduct a preliminary exploratory data analysis (EDA) of the tanker trip emissions data. We will examine the distribution of emissions across voyages, identify key ports and routes, and explore potential trends in the data.


First we look at the total amount of CO2 emissions, number of tanker trips, and number of unique tanker vessels, and  for all tanker voyages that end at California ports in our dataset, aggregated by tanker type. The class with by far the most activity is `tanker.chemical_oil`, which includes crude oil tankers, petroleum product tankers, and chemical tankers. We cannot at this time differentiate further between chemica and oil tankers, but we should eventually have this capability later this year. `tanker.liquified_gas` are much less common in voyages ending in California.

For the remainder of the analysis after this figure, we will focus on the `tanker.chemical_oil` class.

```{r}
#| label: fig-vessel-class-summary
#| fig-cap: "Summary of CO2 emissions, tanker trips, and unique tanker vessels, by vessel class, for voyages ending at California ports."
#| fig-height: 3
tanker_trip_emissions_ending_in_california |>
  dplyr::group_by(vessel_class) |>
  dplyr::summarise(
    total_trips = dplyr::n_distinct(trip_id),
    total_vessels = dplyr::n_distinct(ssvid),
    emissions_co2_mt = sum(emissions_co2_mt)
  ) |>

  dplyr::arrange(-total_trips) |>
  tidyr::pivot_longer(
    cols = c(total_trips, total_vessels, emissions_co2_mt),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      "emissions_co2_mt" = "CO2 emissions (MT)",
      "total_trips" = "Total trips",
      "total_vessels" = "Total vessels"
    )
  ) |>
  dplyr::mutate(
    class_of_interest = ifelse(
      vessel_class == "tanker.chemical_oil",
      TRUE,
      FALSE
    )
  ) |>
  ggplot(aes(
    y = reorder(vessel_class, value),
    x = value,
    fill = class_of_interest
  )) +
  geom_bar(stat = "identity") +
  facet_wrap(~metric, scales = "free_x") +
  theme_minimal() +
  labs(x = "", y = "Vessel class") +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set1", direction = -1)

```


```{r}
#| label: data-country-summary

countries_of_interest <- c("South Korea", "India")
vessel_class_of_interest <- "tanker.chemical_oil"

country_level_data <- tanker_trip_emissions_ending_in_california |>
  dplyr::filter(vessel_class == vessel_class_of_interest) |>
  dplyr::group_by(from_country_iso3) |>
  dplyr::summarise(
    total_trips = dplyr::n_distinct(trip_id),
    total_vessels = dplyr::n_distinct(ssvid),
    emissions_co2_mt = sum(emissions_co2_mt)
  ) |>
  dplyr::arrange(-total_trips) |>
  dplyr::mutate(
    from_country = countrycode::countrycode(
      from_country_iso3,
      "iso3c",
      "country.name",
      nomatch = from_country_iso3
    )
  ) |>
  dplyr::mutate(
    country_of_interest = ifelse(
      from_country %in% countries_of_interest,
      TRUE,
      FALSE
    )
  ) |>
  dplyr::mutate(rank = dplyr::row_number())
```

Next we look at the total amount of CO2 emissions, number of tanker trips, and number of unique tanker vessels, and  for all chemical and oil tanker voyages that end at California ports in our dataset, aggregated by departure country. In terms of CO2 emissions, our countries of interest, `r country_level_data |> dplyr::filter(country_of_interest) |> dplyr::pull(from_country) |> paste0(collapse = " and ")`, rank `r country_level_data |> dplyr::filter(country_of_interest) |> dplyr::pull(rank) |> paste0(collapse = " and ")` respectively out of `r nrow(country_level_data)` countries with voyages ending in California.

For the remainder of the analysis after this figure, we will focus on voyages departing from our countries of interest, `r country_level_data |> dplyr::filter(country_of_interest) |> dplyr::pull(from_country) |> paste0(collapse = " and ")`.

```{r}
#| label: fig-country-summary
#| fig-cap: "Summary of CO2 emissions, tanker trips, and unique tanker vessels, by departure country, for voyages ending at California ports."
#| fig-height: 10
country_level_data |>
  tidyr::pivot_longer(
    cols = c(total_trips, total_vessels, emissions_co2_mt),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      "emissions_co2_mt" = "CO2 emissions (MT)",
      "total_trips" = "Total trips",
      "total_vessels" = "Total vessels"
    )
  ) |>
  ggplot(aes(
    y = reorder(from_country, value),
    x = value,
    fill = country_of_interest
  )) +
  geom_bar(stat = "identity") +
  facet_wrap(~metric, scales = "free_x") +
  theme_minimal() +
  labs(x = "", y = "Departure country") +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set1", direction = -1)

```